---
title: "Weather Event Analysis"
author: "DBisure"
date: "03/07/2020"
output: html_document
---


# Synopsis
There are many weather events happening throughout the world. Some events can be destructive not only to human health but can causive massive economic damage. In this analysis we will try to figure out which weather events have the most effect. To make this analyis we wil be using the dataset from National Weather Service. This dataset contains data from 1950 to end of November 2011.

# Data Processing
We will download the file using R. We will summarizes the data and check for any missing or NA values

```{r}
if (!file.exists("StormData.csv.bz2")){ 
       fileURL <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'
       download.file(fileURL, destfile='StormData.csv.bz2')
}
storm <- read.csv(bzfile('StormData.csv.bz2'),header=TRUE, stringsAsFactors = FALSE)
summary(storm)
```

We are concerned with only a few columns namely: EVTYPE, FATALITIES, INJURIES, PROPDMG, CROPDMG.
Lets see if there are any NA values present

```{r}
mean(is.na(storm$EVTYPE))
mean(is.na(storm$FATALITIES))
mean(is.na(storm$INJURIES))
mean(is.na(storm$PROPDMG))
mean(is.na(storm$CROPDMG))
```
No column consists of NA values

# Results
We will addressing 2 Questions, which weather event causes maximum human damage and which event causes maximum economic damage

## Finding Total Human Damage

### Total Fatalities
```{r}
FatalityData <- aggregate(storm$FATALITIES, by = list(storm$EVTYPE), FUN = sum)
names(FatalityData) <- c("Event","Fatalities")
FatalityData <- FatalityData[order(-FatalityData$Fatalities),]
FatalityData <- FatalityData[1:20,]
head(FatalityData)
```

### Total Injuries

```{r}
InjuryData <- aggregate(storm$INJURIES, by = list(storm$EVTYPE), FUN = sum)
names(InjuryData) <- c("Event","Injuries")
InjuryData <- InjuryData[order(-InjuryData$Injuries),]
InjuryData <- InjuryData[1:20,]
head(InjuryData)

```

### Creating Plot

```{r}
library(ggplot2)

fatalplot <- ggplot(FatalityData,aes(y=Fatalities,x= reorder(Event, -Fatalities))) +
 geom_bar(stat="identity",fill="steelblue") +
 theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5)) +
 ggtitle("Events with Highest Fatality") + xlab("Events") + ylab("Number of Fatalities")

injuryplot <- ggplot(InjuryData,aes(y=Injuries,x= reorder(Event, -Injuries))) +
  geom_bar(stat="identity",fill="red") +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5)) +
  ggtitle("Events with Highest Injuries") + xlab("Events") + ylab("Number of Injuries")

library(gridExtra)
grid.arrange(fatalplot,injuryplot,ncol=2)
```

### As it can be seen Tornado causes maximum fatalities as well injuries

## Finding Total Economic Damage

### Total Property Damage and Total Crop Damange
```{r}
propdamage <- aggregate(storm$PROPDMG, by = list(storm$EVTYPE),FUN=sum)
names(propdamage) <- c("Event","PropDamage")
cropdamage <- aggregate(storm$CROPDMG, by = list(storm$EVTYPE),FUN=sum)
names(cropdamage) <- c("Event","CropDamage")
TotalDamage <- merge(propdamage,cropdamage, by.x = "Event", by.y = "Event")
TotalDamage$Total <- TotalDamage$PropDamage + TotalDamage$CropDamage
TotalDamage <- TotalDamage[order(-TotalDamage$Total),]
TotalDamage <- TotalDamage[1:20,]
head(TotalDamage,n=10)
```

### Creating a plot 

```{r}
options(scipen = 3500000)
damageplot <- ggplot(TotalDamage, aes(x=reorder(Event, -Total),y = Total)) +                          
  geom_bar(stat="identity",fill="red") +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5)) +
  ggtitle("Events with Highest Damage") + xlab("Events") + ylab("Amount of Damage($)")
damageplot
```

### As we can see Tornadoes cause maximum economic damage